---
title: "p8105_hw1_kp3127"
output: html_document
date: "2025-09-19"
---

#problem 1
```{r}
library("tidyverse")
library("moderndive") 
data("early_january_weather")
names("early_january_weather")
```

The `early_january_weather` dataset contains `r nrow(early_january_weather)` rows and `r ncol(early_january_weather)` columns of hourly weather observations.  
Key variables include:  
**time_hour**, which records the date and hour of each 
measurement (the first entry is `r early_january_weather$time_hour[1]`).  
**temp**, the air temperature in degrees Fahrenheit with a mean of 
`r round(mean(early_january_weather$temp, na.rm = TRUE), 1)` °F.  
**humid**, the relative humidity with a mean of `r round(mean(early_january_weather$humid, na.rm = TRUE), 2)`.  
Other variables provide information on wind speed, precipitation, etc.

```{r echo=FALSE}
ggplot(early_january_weather, aes(x = time_hour, y = temp, color = humid)) +
  geom_point()
```

```{r echo=FALSE}
ggsave("temp_vs_time.png")
```

#problem 2  
```{r}
set.seed(1)

df = tibble(
  norm_samp = rnorm(10),                     # numeric
  greater0 = norm_samp > 0,                  # logical
  char_vec = letters[1:10],                  # character
  factor_vec = factor(rep(c("low","med","high"), length.out = 10)) # factor
)
```

```{r}
mean(pull(df,norm_samp))      # works (numeric)
mean(pull(df,greater0))       # works (TRUE=1, FALSE=0)
mean(pull(df,char_vec))       # error
mean(pull(df,factor_vec))     # error
```

```{r}
as.numeric(pull(df, greater0))   # logical -> 0 (FALSE) / 1 (TRUE)
as.numeric(pull(df, char_vec))   # character -> NA with warning
as.numeric(pull(df, factor_vec)) # factor -> underlying integer codes (1, 2, 3)
```
Converting variables with `as.numeric` shows why the `mean()` function succeeds or fails.  
Logical values become 0/1, so their mean represents the proportion of `TRUE`s. 
Character values cannot be converted to numbers, so the mean fails.   
Factors are stored internally as integer codes, so taking their mean would average those codes, which is not meaningful for categorical data.   
This explains why only the numeric and logical variables produce interpretable means.